---
title: "robservable Gallery"
author: "Julien Barnier, Kenton Russel"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 5
    toc: true
vignette: >
  %\VignetteIndexEntry{robservable Gallery}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---



```{r, include=FALSE}
library(robservable)
library(htmlwidgets)
knitr::opts_chunk$set(
  screenshot.force = FALSE,
  echo = TRUE
)
```


The goal of this vignette is to show some examples of (hopefully) useful, interesting or fun notebooks usable with `robservable`.S

## Bivariate Choropleth

The following notebook allows to generate bivariate choropleth maps with zoom and tooltips :

https://observablehq.com/@juba/reusable-bivariate-choropleth

We first load some data from the [USA.county.data Github project](https://github.com/Deleetdk/USA.county.data), recode identifiers and select two of the available variables :

```{r}
load(url("https://raw.githubusercontent.com/Deleetdk/USA.county.data/master/data/USA_county_data.RData"))

d <- USA_county_data 
d$fips = stringr::str_pad(d$fips, 5, side="left", pad = "0")
d <- d[, c("fips", "name_16", "Graduate.Degree", "Less.Than.High.School")]
names(d) <- c("fips", "name_16", "Graduate", "<High.School")
```

Then we can call `robservable` to load the notebook, render only `chart` and `draw` (both are needed for the map to show up), hide `draw` and update a bunch of cells values via the `input` named list. You can refer to the notebook for an explanation of the different values.

```{r}
robservable(
  "@juba/reusable-bivariate-choropleth",
  include = c("chart", "draw"),
  hide = "draw",
  input = list(
    data = d,
    data_id = "fips",
    data_name = "name_16",
    data_var1 = "Graduate",
    data_var2 = "<High.School",
    map = "https://raw.githubusercontent.com/jgoodall/us-maps/master/topojson/county.topo.json",
    map_object = "county",
    map_id_property = NULL,
    contour = "https://raw.githubusercontent.com/jgoodall/us-maps/master/topojson/state.topo.json",
    contour_object = "state",
    contour_width = 2,
    projection = "geoAlbersUsa",
    legend_position = "bottomleft"
  ),
  width = 700,
  height = 700
)
```





