---
title: "robservable in Shiny applications"
author: "Julien Barnier, Kenton Russell"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 5
    toc: true
vignette: >
  %\VignetteIndexEntry{robservable in Shiny applications}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---



```{r, include=FALSE}
library(robservable)
library(htmlwidgets)
knitr::opts_chunk$set(
  screenshot.force = FALSE,
  echo = TRUE
)
```


## Basic usage

`robservable` can be used inside a Shiny app the same way as any `htmlwidget`. Widgets are placed in the `ui` with `robservableOutput`, and are rendered in `server` with `renderRobservable`.

You will find the following basic example in [examples/01_shiny_basic.R](https://github.com/juba/robservable/blob/master/examples/01_shiny_basic.R).


```{r, eval = FALSE}
## Basic robservable Shiny app

library(shiny)
library(robservable)

data("USPersonalExpenditure")

ui <- fluidPage(
    titlePanel("robservable basic Shiny app"),
    sidebarLayout(
        sidebarPanel(
            selectInput(
                "year",
                label = "Year",
                choices = colnames(USPersonalExpenditure)
            )
        ),
        mainPanel(
           robservableOutput("chart", width = 600)
        )
    )
)


server <- function(input, output) {

    df <- reactive({
        data.frame(
            name = rownames(USPersonalExpenditure),
            value = USPersonalExpenditure[, input$year]
        )
    })

    output$chart <- renderRobservable({
        robservable(
            "@d3/horizontal-bar-chart",
            include = "chart",
            input = list(
                data = df(),
                margin = list(top = 20, right = 20, bottom = 30, left = 130)
            ),
            width = 600
        )
    })
}

shinyApp(ui = ui, server = server)
```

As for any `robservable` usage, we have to ensure that the data passed to the widget is in the same format as the one used in the notebook (same shape, same column names).

Another example displaying a multi-line chart with some interactivity can be found in [examples/02_shiny_simple.R](https://github.com/juba/robservable/blob/master/examples/02_shiny_simple.R).


## Animated transitions

When the notebook is designed appropriately, charts can be updated with animated transitions instead of being recreated from scratch. You will find an example of these in [examples/03_shiny_updatable.R](https://github.com/juba/robservable/blob/master/examples/03_shiny_updatable.R), and the [associated notebook](https://observablehq.com/@juba/updatable-bar-chart).

For more examples, the following notebooks provide charts that are updatable with animated transitions and who follow the pattern given at the end of the [Learn D3 animation](https://observablehq.com/@d3/learn-d3-animation#chart) notebook :

- [Updatable bar chart](https://observablehq.com/@juba/updatable-bar-chart)
- [Draggable Pie/Donut chart](https://observablehq.com/@juba/draggable-pie-donut-chart)
- [Reusable bivariate choropleth map](https://observablehq.com/@juba/reusable-bivariate-choropleth)
- [Stacked-to-Grouped Bars](https://observablehq.com/@d3/stacked-to-grouped-bars)


## Observers -- using robservable widgets as inputs

You can add observers to an robservable widget and use its associated cell values as inputs inside your Shiny app. 

This can allow, for example, to use any input defined in the [famous Inputs notebook](https://observablehq.com/@jashkenas/inputs) to be used as an input in a Shiny app.

In the example in [examples/04_shiny_observers.R](https://github.com/juba/robservable/blob/master/examples/04_shiny_observers.R), we use [worldMapCoordinates](https://observablehq.com/@jashkenas/inputs#worldMapCoordinates) as an input to update a leaflet map.


To add an observer to an included cell, you must add the cell names to the `observers` argument.

```r
robservable(
    "@jashkenas/inputs",
    include = c("worldMapCoordinates", "viewof worldMap1"),
    hide = "worldMapCoordinates",
    observers = "worldMap1"
)
```

The cell values will then be accessible via the Shiny `input` object by prefixing the observer name with the `robservable` Shiny id. So, for example, you can use the previous `robservable` instance as an `robservableOutput` with id `map_input`.

```r
## In ui
robservableOutput("map_input")

## In server
output$map_input <- renderRobservable({
    robservable(
        "@jashkenas/inputs",
        include = c("worldMapCoordinates", "viewof worldMap1"),
        hide = "worldMapCoordinates",
        observers = "worldMap1"
    )
})
```

Then the values will be accessible in `input$map_input_worldMap1`.

You can use a named list as `observers` argument to specify custom Shiny `input` names, but you will still have to add the `id` (`map_input` in this example) prefix. So with the following, values will be accessible in `input$map_input_point`.

```r
robservable(
    "@jashkenas/inputs",
    include = c("worldMapCoordinates", "viewof worldMap1"),
    hide = "worldMapCoordinates",
    observers = list(point = "worldMap1")
)
```

## Proxy usage

`robservable` is built to update rather than completely re-render any `renderRobservable` calls after initial creation.  However the updating `renderRobservable` command might need to contain all of the arguments used in the original render.  `robservable` provides proxy methods to ease manipulation of the widget once it has been rendered.  Currently, the package supports the following proxy methods:

- `robs_update` to update observable inputs or variables from Shiny
- `robs_observe` to add an observer on an observable input or variable that communicates with Shiny.

You can find examples of these proxy methods usage in several sample applications :

- [proxy_simple](https://github.com/juba/robservable/blob/master/examples/05_proxy_simple.R) : simple application which modifies color, height and width of a chart via `robs_update`.
- [proxy_data](https://github.com/juba/robservable/blob/master/examples/06_proxy_data.R) : same application but with chart data updating.
- [proxy_updatable](https://github.com/juba/robservable/blob/master/examples/07_proxy_updatable.R) : another chart with data updating via `robs_update`, but with animated transitions.
- [proxy_observer](https://github.com/juba/robservable/blob/master/examples/08_proxy_observer.R) :  sample application which sets a cell value observer with `robs_observe`.





